---
title: "joining table"
format: html

execute:
  cache: true

---

In this document, we'll explore with very simple and small tables how to use the different `join` functions to merge tables together.

### Packages

First, as usual, we need to load the `tidyverse` package.

```{r}
library(tidyverse)
```

We also load the `tidylog` package which is very useful for beginners as, for most `dplyr` and `tidyr` functions, it adds a message to the console explaining in plain English what the function did.

```{r}
library(tidylog)
```

We also load the `pander` package to print the tables in a pretty way:

```{r}
library(pander)
```

### Joining two tables

We create two simple tables that contain two columns: a `sample` column providing the sample ID, and a `value` column that contains the corresponding values.

```{r}

A <- tibble(sample = c("a","b","c"), value_A = c(1,2,3))
A |> pander()

B <- tibble(sample = c("a","b"), value_B = c(4,5))
B |> pander()
```

```{r}

A |> left_join(B) |> pander()
B |> left_join(A) |> pander()

```

```{r}

A |> full_join(B) |> pander()

```

We see that, automatically, `left_join` and `full_join` identify the column that exist in both tables and join using that column.

### Joining when no common column

If we do not have any column in common, we get an error. Let's try introducing another small table:

```{r}

C <- tibble(sid = c("a","b","c"), value_C = c(7,8,9))
C |> pander()

```

```{r}
#| error: true
A |> left_join(C) |> pander()

```

If we know that these two tables are, in fact, reporting values for the same samples, we can rename the `sid` column so that it matches the name in table A

```{r}

C2 <- C |> rename(sample = sid)
C2  |> pander()

A |> left_join(C2) |> pander()

```

### Joining when no samples in common

Now, let's try to join with a table that does not have any sample in common with A:

```{r}

D <- tibble(sample = c("d","e","f"), value_D = c(1, 11, 111))
D |> pander()

```

```{r}

A |> left_join(D) |> pander()

A |> full_join(D) |> pander()

A |> inner_join(D) |> pander()

```

### How do we reorder the columns from the joined table?

For example, if we want to see the `value_B` column first after joining A and B, we can do:

```{r}

A |>
  left_join(B) |> 
  select(value_B, everything()) |> pander()

```

What if there are **several rows** for each sample in a table we want to join with?

```{r}

G <- tibble(sample = rep(c("a","b","c"), each = 2), timepoint = rep(c("1","2"),3), value_G = 1:6)
G |> pander()

```

```{r}

A |> left_join(G) |> pander()
G |> left_join(A) |> pander()


A |> full_join(G) |> pander()

A |> inner_join(G) |> pander()

```

### Pivots

How do we transform table G so that we have only one row per sample and two columns: one for each time-point?

```{r}

G_wide <- 
G |> 
  pivot_wider(
   id_cols = sample,
   names_from = timepoint,
   names_prefix = "time_point",
   values_from = value_G
  ) 

G_wide |> pander()



```

And we can go back to the table in long format by calling the `pivot_longer` function:

```{r}

G_wide |> 
  pivot_longer(
    cols = starts_with("time_point"),
    names_to = "timepoint",
    names_pattern = "time_point(.)",
    values_to = "x_g"
  ) |> pander()
```
