

### Coding exercise 7.1

In this worksheet, we will discuss how to change and customize color scales.

We will be using the R package **tidyverse**, which includes `ggplot()` and related functions. We will also be using the R package **colorspace** for the scale functions it provides.

```{r library-calls}
#| eval: TRUE
#| echo: TRUE
#| output: FALSE

# load required library
library(tidyverse)
library(colorspace)

temperatures <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  mutate(
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(location, day_of_year, month, temperature)

temps_months <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  group_by(location, month_name) %>%
  summarize(mean = mean(temperature)) %>%
  mutate(
    month = factor(
      month_name,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    ),
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(-month_name)
```

We will be working with the dataset `temperatures` that we have used in previous worksheets. This dataset contains the average temperature for each day of the year for four different locations.

```{r temperatures}
#| eval: TRUE
#| echo: TRUE

temperatures
```

We will also be working with an aggregated version of this dataset called `temps_months`, which contains the mean temperature for each month for the same locations. 

```{r temps_months}
#| eval: TRUE
#| echo: TRUE
temps_months
```

As a challenge, try to create this above table yourself using `group_by()` and `summarize()` like we learned about Wednesday., and then make a `month` column which is a factor with levels froing from "Jan" to "Dec", and make the location column a factor with levels "Death Valley", "Houston", "San Diego", "Chicago".  If you are having trouble, the solution is at the end of this page, make sure you copy it into your code so the rest of the exercise works.


```{r temp_months_by_user}
#| eval: FALSE
#| echo: TRUE

# check solution at the end before moving on!
temps_months <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  group_by(___) %>%
  summarize(___) %>%
  mutate(
    month = factor(
      month_name,
      ___
    ),
    location = factor(
      location, ___
    )
  ) %>%
  select(-month_name)
```

#### Built in **ggplot2** color scales

We will start with built-in **ggplot2** color scales, which require no additional packages. The scale functions are always named `scale_color_*()` or `scale_fill_*()`, depending on whether they apply to the `color` or `fill` aesthetic. The `*` indicates some other words specifying the type of the scale, for example `scale_color_brewer()` or `scale_color_distiller()` for discrete or continuous scales from the ColorBrewer project, respectively. You can find all available built-in scales here: https://ggplot2.tidyverse.org/reference/index.html#section-scales

Now consider the following plot.

```{r temps-months-default}
#| eval: TRUE
#| echo: TRUE
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile() + 
  coord_fixed(expand = FALSE)
```

If you wanted to change the color scale to one from the ColorBrewer project, which scale function would you have to add?
`scale_color_brewer()`, `scale_color_distiller()`, `scale_fill_brewer()`, `scale_fill_distiller()`?

```{r temps-months-default1}
 # answer the question above to yourself
```

Now try this out.

```{r temps-months-distiller}
#| eval: FALSE
#| echo: TRUE
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile() + 
  coord_fixed(expand = FALSE) +
  ___
```

Most color scale functions have additional customizations. How to use them depends on the specific scale function. For the ColorBrewer scales you can set `direction = 1` or `direction = -1` to set the direction of the scale (light to dark or dark to light). You can also set the palette via a numeric argument, e.g. `palette = 1`, `palette = 2`, `palette = 3` etc.

Try this out by setting the direction of the scale from light to dark and using palette #4.

```{r temps-months-distiller2}
 # build all the code for this exercise
```

A popular set of scales are the viridis scales, which are provided by `scale_*_viridis_c()` for continuous data and `scale_*_viridis_d()` for discrete data. Change the above plot to use a viridis scale.

```{r temps-months-viridis}
 # build all the code for this exercise
```

The viridis scales can be customized with `direction` (as before), `option` (which can be `"A"`, `"B"`, `"C"`, `"D"`, or `"E"`), and `begin` and `end` which are numerical values between 0 and 1 indicating where in the color scale the data should begin or end. For example, `begin = 0.2` means that the lowest data value is mapped to the 20th percentile in the scale.

Try different choices for `option`, `begin`, and `end` to see how they change the plot.

```{r temps-months-viridis2,}
 # build all the code for this exercise
```
#### Customizing scale title and labels

In a previous worksheet, we used arguments such as `name`, `breaks`, `labels`, and `limits` to customize the axis. For color scales, instead of an axis we have a legend, and we can use the same arguments inside the scale function to customize how the legend looks.

Try this out. Set the scale limits from 10 to 110 and set the name of the scale and the breaks as you wish.

```{r temps-months-viridis3}
#| eval: FALSE
#| echo: TRUE
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile() + 
  coord_fixed(expand = FALSE) +
  scale_fill_viridis_c(
    name = ___,
    breaks = ___,
    limits = ___
  )
```

**Note:** Color scales ignore the `expand` argument, so you cannot use it to expand the scale beyond the data values as you can for position scales.

#### Binned scales

Research into human perception has shown that continuous coloring can be difficult to interpret. Therefore, it is often preferable to use a small number of discrete colors to indicate ranges of data values. You can do this in ggplot with binned scales. For example, `scale_fill_viridis_b()` provides a binned version of the viridis scale. Try this out.

```{r temps-months-viridis4}
#| eval: FALSE
#| echo: TRUE
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile() + 
  coord_fixed(expand = FALSE) +
  ___
```
You can change the number of bins by providing the `n.breaks` argument or alternatively by setting breaks explicitly. Try this out.

```{r temps-months-viridis5}
 # build all the code for this exercise
```
#### Scales from the **colorspace** package

The color scales provided by the **colorspace** package follow a simple naming scheme of the form `scale_<aesthetic>_<datatype>_<colorscale>()`, where `<aesthetic>` is the name of the aesthetic (`fill`, `color`, `colour`), `<datatype>` indicates the type of variable plotted (`discrete`, `continuous`, `binned`), and `colorscale` stands for the type of the color scale (`qualitative`, `sequential`, `diverging`, `divergingx`).

For the mean temperature plot we have been using throughout this worksheet, which 2 color scales from the **colorspace** package is/are appropriate?

`scale_fill_binned_sequential()`, `scale_fill_discrete_qualitative()`, `scale_fill_continuous_sequential()`, `scale_color_continuous_sequential()`, `scale_color_continuous_diverging()`>


```{r temps-months-colorspace}
#| eval: FALSE
#| echo: TRUE
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile() + 
  coord_fixed(expand = FALSE) +
  ___
```
You can customize the colorspace scales with the `palette` argument, which takes the name of a palette (e.g., `"Inferno"`, `"BluYl"`, `"Lajolla"`). Try this out. Also try reversing the scale direction with `rev = TRUE` or `rev = FALSE`. (The **colorspace** scales use `rev` instead of `direction`.) You can find the names of all supported scales [here](https://colorspace.r-forge.r-project.org/articles/hcl_palettes.html) (consider specifically single-hue and multi-hue sequential palettes).


```{r temps-months-colorspace2}
 # build all the code for this exercise
```
You can also use `begin` and `end` just like in the viridis scales.


#### Manual scales

For discrete data with a small number of categories, it's usually best to set colors manually. This can be done with the scale functions `scale_*_manual()`. These functions take an argument `values` that specifies the color values to use.

To see how this works, let's go back to this plot of temperatures over time for four locations.

```{r temps-locations}
#| eval: TRUE
#| echo: TRUE
#| warning: FALSE
ggplot(temperatures, aes(day_of_year, temperature, color = location)) +
  geom_line(size = 1.5)
```
Let's use the following four colors: `"gold2"`, `"firebrick"`, `"blue3"`, `"springgreen4"`. We can visualize this palette using the function `swatchplot()` from the **colorspace** package.


```{r swatchplot}
#| eval: TRUE
#| echo: TRUE
colorspace::swatchplot(c("gold2", "firebrick", "blue3", "springgreen4"))
```
Now apply this color palette to the temperatures plot, by using the manual color scale. *Hint*: use the `values` argument to provide the colors to the manual scale.

```{r temps-locations-manual}
#| eval: FALSE
#| echo: TRUE
ggplot(temperatures, aes(day_of_year, temperature, color = location)) +
  geom_line(size = 1.5) +
  ___
```


One problem with this approach is that we can't easily control which data value gets assigned to which color. What if we wanted San Diego to be shown in green and Chicago in blue? The simplest way to resolve this issue is to use a named vector. A named vector in R is a vector where each value has a name. Named vectors are created by writing `c(name1 = value1, name2 = value2, ...)`. See the following example.

```{r named-vector}
#| eval: TRUE
#| echo: TRUE
# regular vector
c("cat", "mouse", "house")
# named vector
c(A = "cat", B = "mouse", C = "house")
```

The names in the second example are `A`, `B`, and `C`. Notice that the names are not in quotes. However, if you need a name containing a space (such as `Death Valley`), you need to enclose the name in backticks. Thus, our named vector of colors could be written like so:

```{r named-vector2}
#| eval: TRUE
#| echo: TRUE
c(`Death Valley` = "gold2", Houston = "firebrick", Chicago = "blue3", `San Diego` = "springgreen4")
```

Now try to use this color vector in the figure showing temperatures over time.

```{r temps-locations-manual2}
#| eval: FALSE
#| echo: TRUE
 # build all the code for this exercise
```


Try some other colors also. For example, you could use the Okabe-Ito colors:

```{r swatchplot2}
#| eval: TRUE
#| echo: TRUE
# Okabe-Ito colors
colorspace::swatchplot(c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#000000"))
```

Alternatively, you can find a list of all named colors [here.](https://www.datanovia.com/en/blog/awesome-list-of-657-r-color-names/) You can also run the command `colors()` in your R console to get a list of all available color names.

*Hint:* It's a good idea to never use the colors `"red"`, `"green"`, `"blue"`, `"cyan"`, `"magenta"`, `"yellow"`. They are extreme points in the RGB color space and tend to look unnatural and too crazy. Try this by making a swatch plot of these colors, and compare for example to the color scale containing the colors `"firebrick"`, `"springgreen4"`, `"blue3"`, `"turquoise3"`, `"darkorchid2"`, `"gold2"`. Do you see the difference?


```{r swatchplot3}
#| eval: FALSE
#| echo: TRUE
 # build all the code for this exercise
```

*Solution to the challenge to make the summary table of mean temperature by month:*

```{r solution_to_challenge}
#| eval: FALSE
#| echo: TRUE
# paste this below the "temperatures" code-chunk
temps_months <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  group_by(location, month_name) %>%
  summarize(mean = mean(temperature)) %>%
  mutate(
    month = factor(
      month_name,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    ),
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(-month_name)
```