---
title: "Compound Figures Exercise Solutions"
format: 
  html: default


execute:
  cache: true

---



```{r setup3}
#| eval: TRUE
#| echo: FALSE
#| warning: FALSE
library(purrr)
library(tufte)
library(tidyverse)
library(patchwork)
library(cowplot)
library(lubridate)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, eval = FALSE,  echo=TRUE, comment = "", cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)

p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp))

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear))

p3 <- ggplot(mtcars) + 
  geom_smooth(aes(disp, qsec))

p4 <- ggplot(mtcars) + 
  geom_bar(aes(carb))
```

### Coding exercise 7.3

In this worksheet, we will discuss how to combine several **ggplot2** plots into one compound figure.

We will be using the R package **tidyverse**, which includes `ggplot()` and related functions. We will also be using the package **patchwork** for plot composition.

```{r library-calls3}
#| eval: TRUE
#| echo: TRUE
# load required library
library(tidyverse)
library(patchwork)
```

We will be working with the dataset `mtcars`, which contains fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

```{r mtcars}
#| eval: TRUE
#| echo: TRUE
mtcars
```
#### Combining plots

First we set up four different plots that we will subsequently combine. The plots are stored in variables `p1`, `p2`, `p3`, `p4`.

```{r singlet-plot}
#| eval: TRUE
#| echo: TRUE
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp))
p1  

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear))
p2

p3 <- ggplot(mtcars) + 
  geom_smooth(aes(disp, qsec))
p3

p4 <- ggplot(mtcars) + 
  geom_bar(aes(carb))
p4
```


To show plots side-by-side, we use the operator `|`, as in `p1 | p2`. Try this by making a compound plot of plots `p1`, `p2`, `p3` side-by-side.

```{r plots-side-by-side}
#| eval: TRUE
#| echo: TRUE
 # build all the code for this exercisew

  # solution 
 p1 | p2 | p3
```


To show plots on top of one-another, we use the operator `/`, as in `p1 / p2`. Try this by making a compound plot of plots `p1`, `p2`, `p3` on top of each other.

```{r plots-on-top}
#| eval: TRUE
#| echo: TRUE
 # build all the code for this exercise

  p1 / p2 / p3
```


We can also use parentheses to group plots with respect to the operators `|` and `/`. For example, we can place several plots side-by-side and then place this entire row of plots on top of another plot. Try putting `p1`, `p2`, `p3`, on the top row, and `p4` on the bottom row.

```{r plots-complex}
#| eval: FALSE
#| echo: TRUE

(___) / ___
```

```{r plots-complex_solution}
#| eval: FALSE
#| echo: TRUE

# solution
(p1 | p2 | p3 ) / p4
```

#### Plot annotations

The **patchwork** package provides a powerful annotation system via the `plot_annotation()` function that can be added to a plot assembly. For example, we can add plot tags (the labels in the upper left corner identifying the plots) via the plot annotation `tag_levels`. You can set `tag_levels = "A"` to generate tags A, B, C, etc. Try this out.

```{r plots-w-tags}
#| eval: TRUE
#| echo: TRUE
(p1 | p2 | p3 ) / p4
```

```{r plots-w-tags_solution}
#| eval: TRUE
#| echo: TRUE

# solution
(p1 | p2 | p3 ) / p4 +
  plot_annotation(
    tag_levels = "A"
  )
```
Try also tag levels such as `"a"`, `"i"`, or `"1"`.

You can also add elements such as titles, subtitles, and captions, by setting the `title`, `subtitle`, or `caption` argument in `plot_annotation()`. Try this out by adding an overall title to the figure from the previous exercise.

```{r plots-w-title}
#| eval: TRUE
#| echo: TRUE
 # build all the code for this exercise

 # solution
(p1 | p2 | p3 ) / p4 +
  plot_annotation(
    tag_levels = "A",
    title = "Various observations about old cars"
  )
```


Also set a subtitle and a caption.

Finally, you can change the theme of all plots in the plot assembly via the `&` operator, as in `(p1 | p2) & theme_bw()`. Try this out.

```{r themeing}
#| eval: TRUE
#| echo: TRUE
 # build all the code for this exercise

# solution
(p1 | p2) & theme_bw()
```

What happens if you write this expression without parentheses? Do you understand why?


*(Big) Challenge Problem*:

If you have time this morning, or if you want to work on it this afternoon, try analyzing a new dataset to test your R skills. First, learn what the columns mean, what missing values you see, and then start to ask questions about patterns in the data by making plots.   
You can browse the datasets at https://github.com/rfordatascience/tidytuesday/tree/master/data (click on a year folder and go to the README to read about the datasets). Or pick from one of the ones below:

```{r challenge-dataset}
#| eval: TRUE
#| echo: TRUE
# fish consumption in different countries 
consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-12/fish-and-seafood-consumption-per-capita.csv')

# world cup Cricket matches from 1996 to 2005
matches <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-11-30/matches.csv')

# malaria deaths by age across the world and time. 
 malaria_deaths<- readr::read_csv('https://github.com/rfordatascience/tidytuesday/blob/master/data/2018/2018-11-13/malaria_deaths_age.csv')

#meteorites and/or volcanos:
# note to plot a map, try the following:
countries_map <- ggplot2::map_data("world")
world_map<-ggplot() + 
  geom_map(data = countries_map, 
           map = countries_map,aes(x = long, y = lat, map_id = region, group = group),
           fill = "white", color = "black", size = 0.1) # then add geom_point() to this map to add lat/long points

meteorites <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-06-11/meteorites.csv")
volcano <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv")

# or pick any dataset from https://github.com/rfordatascience/tidytuesday/tree/master/data , 
# just click on a year folder and go to the README to read about the datasets
# if you have trouble loading a dataset there, ask for help!
```