---
title: "Figure Design"
subtitle: "<em>remixed from <a href='https://github.com/wilkelab/SDS375'>Claus O. Wilke's SDS375 course</a></em>"
format: qmdrslides-revealjs
editor_options: 
  chunk_output_type: inline

execute:
  cache: true
---


```{r}
#| echo: FALSE

library(tidyverse)
library(cowplot)
library(ggthemes)
library(colorspace)
library(ggridges)
library(palmerpenguins)
library(here)
# ggplot2 settings so plots scale well for slide size 
#theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

lincoln_temps <- readRDS(here("materials/1-workshop1/7-custom-data-visualizations/slides_files/datasets", "lincoln_temps.rds"))
```

## 
::: {style="font-size: 80%;" layout-ncol=2}

**How to go from this ...**

**... to this?**
```{r}
#| echo: FALSE
ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges()
```




```{r}
#| echo: FALSE
ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#56B4E9", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)#, breaks = c(0, 25, 50, 75)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_minimal_grid() +
  theme( 
    axis.text.y = element_text(vjust = 0)
  )
```

Requires coordinated modification of multiple elements:
\
- geoms (via arguments to geoms)  
- scales (via `scale_*()` functions)  
- plot appearance (via themes)  

:::



## The starting point, a rough draft
\

::: {style="font-size: 70%;" layout-ncol=2}

```{r lincoln-ridgeline-progression1}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges()
```

```{r lincoln-ridgeline-progression1}
#| eval: TRUE
#| echo: FALSE

```

You can download the dataset using this code:

\

```{r}
#| eval: FALSE
#| echo: TRUE
lincoln_temps <- readRDS(
  url("https://wilkelab.org/DSC385/datasets/lincoln_temps.rds")
)
```
:::


## Set `scale` and `bandwidth` to shape ridgelines
\

::: {style="font-size: 70%;" layout-ncol=2}

```{r lincoln-ridgeline-progression2}
#| eval: FALSE
#| echo: TRUE

ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4
  )
```


```{r lincoln-ridgeline-progression2}
#| eval: TRUE
#| echo: FALSE
```
:::


## Set `rel_min_height` to cut ridgelines near zero
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r lincoln-ridgeline-progression3}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  )
```

```{r lincoln-ridgeline-progression3}
#| eval: TRUE
#| echo: FALSE

```
:::

## Use `scale_*()` functions to specify axis labels
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r lincoln-ridgeline-progression4}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)"
  ) +
  scale_y_discrete(
    name = NULL  # NULL means no label
  )
```


```{r lincoln-ridgeline-progression4}
#| eval: TRUE
#| echo: FALSE

```
:::

## Specify scale expansion
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r lincoln-ridgeline-progression5}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  )
```
```{r lincoln-ridgeline-progression5}
#| eval: TRUE
#| echo: FALSE

```

:::

## Set overall plot theme
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r lincoln-ridgeline-progression6}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid()  # from cowplot
```

```{r lincoln-ridgeline-progression6}
#| eval: TRUE
#| echo: FALSE

```


:::

## Align y axis labels to grid lines
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r lincoln-ridgeline-progression7}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```

```{r lincoln-ridgeline-progression7}
#| eval: TRUE
#| echo: FALSE

```
:::

## Change fill color from default gray to blue
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r lincoln-ridgeline-progression8}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
    fill = "#7DCCFF"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
```{r lincoln-ridgeline-progression8}
#| eval: TRUE
#| echo: FALSE

```
:::



## Draw lines in white instead of black
\

::: {style="font-size: 70%;" layout-ncol=2}

```{r lincoln-ridgeline-final}
#| eval: FALSE
#| echo: TRUE
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
    fill = "#7DCCFF",
    color = "white"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```

```{r lincoln-ridgeline-final}
#| eval: TRUE
#| echo: FALSE
```

:::



# Working with ggplot themes
\


## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r penguins-complete-themes}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point()
# default theme is theme_gray()
```
```{r penguins-complete-themes}
#| eval: TRUE
#| echo: FALSE

```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_gray()
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_gray(14) # most themes take a font-size argument to scale text size
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_bw(14)
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_minimal(14)
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_classic(14)
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_half_open()  # from package cowplot
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_minimal_grid()  # from package cowplot
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_minimal_hgrid()  # from package cowplot
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_minimal_vgrid()  # from package cowplot
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_economist(14)       # from package ggthemes
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_economist(14) + scale_color_economist() # from package ggthemes
```
:::

## Using ready-made themes
\

::: {style="font-size: 70%;"}
```{r}
#| eval: TRUE
#| echo: TRUE
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() + 
  theme_fivethirtyeight(14) + scale_color_fivethirtyeight() # from package ggthemes
```
:::


## Customizing theme elements
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements1}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid()
```
```{r penguins-theme-elements1}
#| eval: TRUE
#| echo: FALSE

```
:::

## Customizing theme elements
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements2}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change overall font family
    # (requires font to be available)
    text = element_text(
      family = "Comic Sans MS"
    )
  )
```
```{r penguins-theme-elements2}
#| eval: TRUE
#| echo: FALSE

```
:::

## Customizing theme elements
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements3}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change color of axis titles
    axis.title = element_text(
      color = "royalblue2"
    )
  )
```
```{r penguins-theme-elements3}
#| eval: TRUE
#| echo: FALSE

```
:::

## Customizing theme elements
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements4}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change color of only the x axis title
    axis.title.x = element_text(
      color = "royalblue2"
    )
  )
```
```{r penguins-theme-elements4}
#| eval: TRUE
#| echo: FALSE

```
:::

## Customizing theme elements
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements5}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change all text colors?
    # why does it not work?
    text = element_text(color = "royalblue2")
  )
```
```{r penguins-theme-elements5}
#| eval: TRUE
#| echo: FALSE

```
:::

## Customizing theme elements
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements6}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    text = element_text(color = "royalblue2"),
    axis.text = element_text(
      color = "royalblue2"
    )
  )
```
```{r penguins-theme-elements6}
#| eval: TRUE
#| echo: FALSE

```
:::

## Customizing theme elements
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements6}
#| eval: FALSE
#| echo: TRUE
```
```{r penguins-theme-elements6}
#| eval: TRUE
#| echo: FALSE

```
:::

The element `axis.text` has its own color set in the theme. Therefore it doesn't inherit from `text`.


## Horizontal and vertical alignment
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements7}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.title.x = element_text(
      # horizontal justification
      # (0 = left)
      hjust = 0
    )
  )
```
```{r penguins-theme-elements7}
#| eval: TRUE
#| echo: FALSE
```
:::

## Horizontal and vertical alignment
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements8}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.title.x = element_text(
      # horizontal justification
      # (0.5 = center)
      hjust = 0.5
    )
  )
```
```{r penguins-theme-elements8}
#| eval: TRUE
#| echo: FALSE
```
:::

## Horizontal and vertical alignment
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements9}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.title.x = element_text(
      # horizontal justification
      # (1 = right)
      hjust = 1
    )
  )
```
```{r penguins-theme-elements9}
#| eval: TRUE
#| echo: FALSE
```
:::

## Horizontal and vertical alignment
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements10}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(
      # vertical justification
      # (0 = bottom)
      vjust = 0
    )
  )
```
```{r penguins-theme-elements10}
#| eval: TRUE
#| echo: FALSE
```
:::

## Horizontal and vertical alignment
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements11}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(
      # vertical justification
      # (0.5 = center)
      vjust = 0.5
    )
  )
```
```{r penguins-theme-elements11}
#| eval: TRUE
#| echo: FALSE
```
:::

## Horizontal and vertical alignment
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements12}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) + 
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(
      # vertical justification
      # (1 = top)
      vjust = 1
    )
  )
```
```{r penguins-theme-elements12}
#| eval: TRUE
#| echo: FALSE
```
:::

## Remove elements entirely: `element_blank()`


::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements13}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # all text gone
    axis.text = element_blank()
  )
```
```{r penguins-theme-elements13}
#| eval: TRUE
#| echo: FALSE
#| 
```
:::


## Remove elements entirely: `element_blank()`
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements14}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # no axis titles
    axis.title = element_blank()
  )
```
```{r penguins-theme-elements14}
#| eval: TRUE
#| echo: FALSE
```
:::

## Set background color: `element_rect()`
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements15}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    plot.background = element_rect(
      fill = "aliceblue"
    )
  )
```
```{r penguins-theme-elements15}
#| eval: TRUE
#| echo: FALSE
```
:::

## Set background color: `element_rect()`
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements16}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    panel.background = element_rect(
      fill = "aliceblue"
    )
  )
```
```{r penguins-theme-elements16}
#| eval: TRUE
#| echo: FALSE
```
:::

## Set background color: `element_rect()`
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements17}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    legend.box.background = element_rect(
      fill = "aliceblue",
      color = "steelblue4" # line color
    )
  )
```
```{r penguins-theme-elements17}
#| eval: TRUE
#| echo: FALSE
```
:::

## Set background color: `element_rect()`
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements18}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    legend.box.background = element_rect(
      fill = "aliceblue",
      color = "steelblue4" # line color
    ),
    legend.box.margin = margin(7, 7, 7, 7)
  )
```
```{r penguins-theme-elements18}
#| eval: TRUE
#| echo: FALSE
```
:::

## Move the legend: `legend.position`
\

::: {style="font-size: 70%;" layout-ncol=2}
```{r penguins-theme-elements19}
#| eval: FALSE
#| echo: TRUE
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    legend.box.background = element_rect(
      fill = "aliceblue", 
      color = "steelblue4" # line color
    ),
    legend.box.margin = margin(7, 7, 7, 7),
    # relative position inside plot panel
    legend.position = c(1, 0),
    # justification relative to position
    legend.justification = c(1, 0)
  )
```
```{r penguins-theme-elements19}
#| eval: TRUE
#| echo: FALSE
```
:::


# Exercise 
Go to [https://elsherbini.github.io/durban-data-science-for-biology/materials/1-workshop1/7-custom-data-visualizations/#coding-exercise-7.2](https://elsherbini.github.io/durban-data-science-for-biology/materials/1-workshop1/7-custom-data-visualizations/#coding-exercise-7.2) and complete the Compound Figures exercise.
